How to select predictive models for decision-making or causal inference?

### 2025-02-07 how_to_select_latex.tex
prepare main_submission latex and pdf

%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
\author[1,\authfn{1}]{Matthieu Doutreligne}
\author[2]{Gaël Varoquaux}

@@

We evaluate the following causal metrics, oracle and feasible
versions, presented in Table
\ref{tab:evaluation_metrics}:\\
$\widehat{\mu\text{-risk}}_{IPW}^*$,
$\widehat{R\text{-risk}}^*$,
$\widehat{U\text{-risk}}^*$,
$\widehat{\tau\text{-risk}_{IPW}}^*$,
$\widehat{\mu\text{-risk}}$,
$\widehat{\mu\text{-risk}}_{IPW}$,
$\widehat{R\text{-risk}}$,
$\widehat{U\text{-risk}}$,
$\widehat{\tau\text{-risk}_{IPW}}$.
We benchmark the metrics in a variety of settings:
many different simulated data generation
processes and three semi-simulated datasets \footnote{Scripts for the simulations and the selection procedure are available at
    \url{https://github.com/soda-inria/caussim}.
}.

@@

For these three datasets, the family of candidate estimators are gradient
boosting trees for both the response surfaces and the treatment
\footnote{Scikit-learn regressor,
    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
    and classifier,
    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
with S-learner, learning rate
in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
    35, 40\}$ resulting in a family of size 18.

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
feasible causal metrics. We measure this agreement with relative\footnote{To
    remove the variance across datasets (some datasets lead to easier model
    selection than others), we report values for one metric relative to the mean of
    all metrics for a given dataset instance: $\text{Relative} \,
        \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$
(eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neural network, or language
models.

@@

\begin{itemize}
    \item Project name: Caussim
    \item Project home page: \url{https://github.com/soda-inria/caussim}
    \item Operating system(s): Platform independent
    \item Programming language: Python
    \item License: BSD 3-Clause License
\end{itemize}

@@

\begin{itemize}
    \item Project name: Causal Model Selection
    \item Project home page: \url{https://github.com/soda-inria/causal_model_selection}
    \item Operating system(s): Platform independent
    \item Programming language: Python
    \item License: BSD 2-Clause License
\end{itemize}

@@

Detailed explanations on how to generate the datasets are available in the readme of the data section of our github: \url{https://github.com/soda-inria/caussim/tree/main/data}.

\begin{itemize}
    \item ACIC 2016: dataset provided through R package aciccomp2016: \url{https://github.com/vdorie/aciccomp}

    \item ACIC 2018: scaling subset from link to Synapse : \url{https://www.synapse.org/Synapse:syn11738767}

    \item TWINS: dataset from \url{https://github.com/AMLab-Amsterdam/CEVAE}, the version that we are using is on the following zenodo repository: \url{https://zenodo.org/records/14674618}.

    \item Caussim Dataset: generated dataset - instructions from \url{https://github.com/soda-inria/caussim}
\end{itemize}
\subsection{Experiments result data}

The result data generated by the experiments are available at a dedicated zenodo repository \url{https://zenodo.org/records/13765465}. We provide the data for the simulations and the semi-simulated datasets to allow an easy replication of the main graphic (Fig. \ref{fig:relative_kendalls_all_datasets}) in the result section (detailed steps on our github: \url{https://github.com/soda-inria/caussim/tree/main?tab=readme-ov-file#reports}).

@@

\subsubsection{Additional Results}

>>>>

%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
\author[1,\authfn{1}]{Matthieu Doutreligne (ORCID: 0000-0001-8072-9966)}
\author[2]{Gaël Varoquaux (ORCID: 0000-0003-1076-5122)}

@@

We evaluate the following causal metrics, oracle and feasible
versions, presented in Table
\ref{tab:evaluation_metrics}:\\
$\widehat{\mu\text{-risk}}_{IPW}^*$,
$\widehat{R\text{-risk}}^*$,
$\widehat{U\text{-risk}}^*$,
$\widehat{\tau\text{-risk}_{IPW}}^*$,
$\widehat{\mu\text{-risk}}$,
$\widehat{\mu\text{-risk}}_{IPW}$,
$\widehat{R\text{-risk}}$,
$\widehat{U\text{-risk}}$,
$\widehat{\tau\text{-risk}_{IPW}}$.
We benchmark the metrics in a variety of settings:
many different simulated data generation
processes and three semi-simulated datasets.We provide scripts for the simulations \cite{doutreligne2025caussim} and the selection procedure \cite{doutreligne2025causal}.

@@

For these three datasets, the family of candidate estimators are gradient
boosting trees for both the response surfaces and the treatment
(scikit-learn HistGradientBoostingRegressor HistGradientBoostingClassifier  \cite{pedregosa_scikitlearn_2011})
with S-learner, learning rate
in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
    35, 40\}$ resulting in a family of size 18.

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the
different feasible causal metrics. We measure this agreement with relative
Kendall tau $\kappa$ (eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. To
remove the variance across datasets (some datasets lead to easier model
selection than others), we report values for one metric relative to the mean of
all metrics for a given dataset instance: $\text{Relative} \,
    \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
    mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

To facilitate better model selection, we provide Python
code \cite{doutreligne2025causal}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neural network, or language
models.

@@

\begin{itemize}
    \item Project name: Caussim
    \item Project home page: \cite{doutreligne2025caussim}
    \item Operating system(s): Platform independent
    \item Programming language: Python
    \item License: BSD 3-Clause License
\end{itemize}

@@

\begin{itemize}
    \item Project name: Causal Model Selection
    \item Project home page: \cite{doutreligne2025causal}
    \item Operating system(s): Platform independent
    \item Programming language: Python
    \item License: BSD 2-Clause License
\end{itemize}

@@

Detailed explanations on how to generate the datasets are available in the readme of the data section of our github \cite{doutreligne2025caussim}.

\begin{itemize}
    \item ACIC 2016: dataset provided through R package aciccomp2016 \cite{dorie_automated_2019}

    \item ACIC 2018: scaling subset from the Synapse repository : \cite{shimoni2018ibm}

    \item TWINS: dataset from \cite{louizos_causal_2017}, our version is on a zenodo repository \cite{doutreligne2025twins}.

    \item Caussim Dataset: generated dataset with the source code \cite{doutreligne2025caussim}.
\end{itemize}
\subsection{Experiments result data}

The result data generated by the experiments are available at a dedicated zenodo repository \cite{doutreligne2024how}. We provide the data for the simulations and the semi-simulated datasets to allow an easy replication of the main graphic (Fig. \ref{fig:relative_kendalls_all_datasets}) in the result section.

@@

\subsubsection{Additional Results}\label{apd:experiments:additional_results}
